Transparent Changes

### 2022-05-02 thesis/introduction.Rmd
Continue writing thesis

<!-- VO2max is important -->
<!-- physiological determinants -->
<!-- VO2max testing -->
<!-- Influence factors in testing -->

@@

<!-- First example in 90s: applied research -->
<!-- Since then: various example -->
<!-- conceptual work: Robergs -->
<!-- Comparison: martin-rincon -->
<!-- Problem: group + individual -->
<!-- Knowledge Gap -->

@@

<!-- Gap -> we need this: -->
<!-- overview what is actually done -->
<!-- and how does this impact VO2max assessment -->

<!-- Review: mapping current methodological (+ reporting) state  -->
<!-- Comparison: highlight advantage over previous -->

<!-- help compare results across studies + for individuals -->
<!-- advise for use of methods  -->
<!-- call for improved reporting -->

>>>>

### Maximum oxygen uptake: Definition and Relevance

Performance in endurance sports is limited by the human physiology. Athletes need to supply energy to locomotion, a process that happens mainly via the oxidative phosphorylation. A higher maximal activity of the oxidative phosphorylation allows to supply more energy, and thus to move faster. On a whole-body level the highest activity of oxidative phosphorylation can be approximated by measuring the maximum oxygen uptake (V̇O~2max~).

V̇O~2max~ is defined as the highest rate a body can consume oxygen. Determined by measuring gas exchange data, it is one of the most common measures investigated in sports science and medicine. V̇O~2max~ highly corresponds with endurance performance in heterogeneous groups, and thus can be regarded the most relevant physiological parameter for endurance performance prediction. Many training programs aim to target the V̇O~2max~, and studies evaluate the quality of interventions by measuring changes in V̇O~2max.~

### Measuring the VO~2max~

Researchers measure V̇O~2max~ during exercise tests to exhaustion. In a laboratory setting such exercise commonly takes place on treadmills or cycling Prior fatigue results in underestimating the true V̇O~2max~. Therefore most exercise protocol consist of a continuous or stepwise, fast increase in load. As the oxygen kinetics themselves comprise a time lag, XXX suggested an optimal duration of 8-12 minutes for testing V̇O~2max~. Despite protocol type and duration, factors such as nutrition, XXX bias the V̇O~2max~ determination.

Early research in V̇O~2max~ used bags --- the so called Douglas bag method --- to collect expired air for later analysis. Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chamber (which sample data at fixed time intervals, e.g., 15s) or breath-by-breath methods. While mixing chambers are regarded as more reliable, measuring breath-by-breath generates more data with a higher temporal resolution. <!--Which is more common?--> The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~. Another important, yet often overlooked factor influencing the V̇O~2max~ is the data processing of the raw data.

### VO~2max~ vs VO~2peak~

Whether an observed peak in oxygen uptake corresponds to the true maximum has been extensively discussed in the past decades. To identify a 'true' maximum researchers evaluate a set of parameters measured in a ramp test --- the V̇O~2max~ criteria [@howley1995]. If these criteria a (partly) not fulfilled, researchers should speak of the peak oxygen uptake instead of V̇O~2max~. In this thesis I will not distinguish between peak and maximum oxygen uptake, as the criteria for V̇O~2max~ themselves (e.g. the primary criterium of a plateau in oxygen uptake; the secondary criterium of the maximum respiratory quotient) heavily rely on the data processing strategy used [@astorino2009]. I thus speak of V̇O~2max~ as the maximum oxygen uptake measured during an appropriate exercise test regardless of any V̇O~2max~ criteria.

@@

Measured breath-by-breath data is noisy (see grey points in Figure 1). Both the biological variability of breathing patterns (as well as irregular breaths such as coughs and swallowing) and the measurement error lead to highly fluctuating raw oxygen uptake data. For interpretation the raw data requires some form of processing.

![Raw breath by breath data and different processing strategies during the final minutes of a ramp test to exhaustion. Grey points mark single breaths.](../plots/example.png)

Different data processing strategies influence measured parameters of gas exchange. As soon as when the first automated systems for gas exchange measure were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values. Since then many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that the shorter the calculation interval, the higher the V̇O~2max~ obtained [@johnson1998a; @astorino2009a; @sell2021]. @midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, highlighting that consistency of strategies is key when comparing outcomes. Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear -log equations to compare mean V̇O~2max~ values derived from different processing strategies. There is inconclusive evidence whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

Differences in V̇O~2max~ due to data processing can cause serious implications in practice. They hinder the comparability of data from studies within meta-analysis or the assessment of longitudinal data from athletes that used different labs [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterium of V̇O~2max~ [@astorino2009]. Crucially, in situation where individuals are classified by their V̇O~2max~ --- for example when describing the trainings status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassification [@johnson1998].

Calls to standardize data processing strategies of gas exchange data are frequent (XXX; ...). @howley1995 argued to use longer calculation intervals (60 s), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data @robergs2003a opposed these conclusion, stating that shorter calculation intervals are less erroneous. They argued to use breath-based moving averages as a strategy, as XXX. In a classic review, @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004 . However, @robergs2010 acknowledge the lack of software implementing such procedures.

In absence of one commonly accepted method, data processing varies among the literature. In light of the influence on outcome variables, @myers1990 highlighted the need to to report processing strategies in research articles. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath data in selected journals. They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. While providing interesting first insights into reporting and processing practices, the search by @midgley2007 was not systematic and its methods were not described in detail. @robergs2010 wrote that to investigate the current state of data processing strategies two possible approaches are "(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible". They provides the latter one with a total of 75 respondents, which reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning. Practices in reporting and processing may have changed more than a decade after publication of such numbers.

Astorini 2008: Processing strategy (bbb, 15/30/60 tb) influences VO2max and plateau incidence

Johnson 1998: Processing strategy (bbb, 15/30/60 tb, 8bm) influences VO2max in cardiac patients for heart transplantation evalution

Midgley 2007:

-   processing strategy (tb 10/15/20/30/60) influences VO2max

-   processing strategies vary within the literature (binned time far most common, non-systematic review)

-   processing strategy does not influence reliability of VO2max -\> but consistency is important

Robergs & Burnett 2003: comparing methods (tb, bb, bm) on synthetic and experimental data: Recommendation to use 8 breath moving average. Potential use of filtering methods. Shorter intervals - less error (opposing Howley!)

Howley 1995: classic for VO2max criteria; longer intervals preferable (i.e. 60s), shorter introduce systematic bias towards extreme values (sure???)

Hill 2003: Effect of processing (bm) on VO2max interacts with exercise protocol (= duration)

Schaedler 2017: No interaction between protocol and processing methods (opposing Hill, bit different protocols)

TO READ:

Weir JP, Koerner S, Mack B, Masek J, Vanderhoff J, Heiderscheit BC. VO 2\
plateau detection in cycle ergometry. J Exerc Physiol (2004); 7: 55--62. -\> first with Butterworth (even zero-lag!)

ATS/ACCP Guidelines

Macfarlane DJ. Automated metabolic gas analysis systems: a review. Sports\
Med (2001); 31: 841--861.

Roecker K, Prettin S, Sorichter S. Gas exchange measurements with high\
temporal resolution: the breath-by-breath approach. Int J Sports Med\
(2005); 26(Suppl. 1): S11--S1

<!-- Since then: various example -->

<!-- conceptual work: Robergs -->

<!-- Comparison: martin-rincon -->

<!-- Problem: group + individual -->

<!-- Knowledge Gap -->

@@

<!-- Gap -> we need this: -->

<!-- overview what is actually done -->

<!-- and how does this impact VO2max assessment -->

<!-- Review: mapping current methodological (+ reporting) state  -->

<!-- Comparison: highlight advantage over previous -->

<!-- help compare results across studies + for individuals -->

<!-- advise for use of methods  -->

<!-- call for improved reporting -->

### 2022-05-02 thesis/methods.Rmd
Continue writing thesis

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](). Any deviations from the preregistration are indicated in the 'Transparent Changes' document (LINK/Appendix). Major deviations will also explicitly name be named within the methods section. All data and code of this research project can be found at [GitHub](https://www.github.com/smnnlt/vo2max-processing).

<!-- explicitly state deviations from preregistration -->

@@

The aim of the Scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since the V̇O~2max~ determination is too common to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies was extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.

The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix A contains the corresponding reporting checklist.

@@

```{r terms, echo = FALSE}
out <- data.frame(
  source = c("PubMed", "Web of Science"),
  term = c('((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))', '(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)')
)

@@

The search results from both data bases were joined and checked for the presence of a DOI. Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step. After the removal of duplicates I conducted an automated title scanning to remove results that were likely no original research articles. Table 2 displays the exclusion terms during the automated title screening.

@@

In accordance with the preregistration a random sample was drawn from the search results. The goal of the random sample was to give an unbiased sample of the current state of scientific V̇O~2max~ testing. Based on the procedure described in the preregistration file (LINK), the sample included a total of 500 articles.

The abstracts from the articles included in the random sample were blinded for scanning. This included the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix XXX for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3. When the screeners disagreed in their assessments, conflicts were resolved by discussion.

After the abstract screening I retrieved the full-texts for all articles remaining in the review. The full-texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the full-text exclusion criteria). Conflicts were resolved by discussion.

All data exclusion steps are documented under XXXLINK.

@@

All extracted data is available in the Appendix XXX and under XXXLINK.

@@

The extracted data is presented in a purely descriptive way. I calculated relative and absolute proportions of the different extracted variables.

@@

To determine the influence the most common data processing strategies have on the determination V̇O~2max~, I compared them on a set of already collected gas exchange data from running.

@@

A total of N = 72 exercise test were analysed for this study. The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (XXX male, XXX female; XXX two times). The V̇O~2max~ tests were conducted in XXX to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using the same protocol. Participants run on a treadmill (XXX) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up. After placing the mask for the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

Gas exchange data were recorded using a ZAN XXX device (XXX). The device was calibrated with a 3l-syringe pump (XXX) and a reference gas (XXX) before each measurement. The measured breath-by-breath data is available at XXXLINK.

@@

To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level I calculated the median and 2.5 and 97.5-quantiles of each processing strategy.

>>>>

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s), following the 'Inclusive Systematic Review Registration Form' [@akker2020]. Any deviations from the preregistration are indicated in the 'Transparent Changes' document (Appendix A). Major deviations will also be explicitly stated within the methods section. All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).

@@

The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans.

The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. Appendix B contains the corresponding reporting checklist.

@@

```{r terms, echo = FALSE}
out <- data.frame(
  source = c("PubMed", "Web of Science"),
  term = c('`((((("maximum oxygen uptake") OR ("maximal oxygen uptake")) OR ("VO2max")) OR ("maximum oxygen consumption")) OR ("maximal oxygen consumption")) AND (("2017/01/01"[Date - Publication] : "3000"[Date - Publication]))`', '`(((((ALL=("maximum oxygen uptake")) OR ALL=("maximal oxygen uptake")) OR ALL=("VO2max")) OR ALL=("maximum oxygen consumption")) OR ALL=("maximal oxygen consumption")) AND PY=(2017-2022)`')
)

@@

The search results from both data bases were joined and checked for the presence of a Digital Object Identifier (DOI). Entries without DOI were excluded, to allow for automated removal of duplicates by DOI matching in the next step. After the removing the duplicates I conducted an automated title scanning to exclude results that were likely no original research articles. Table 2 displays the exclusion terms during the automated title screening.

@@

In accordance with the preregistration I drew a random sample from the search results. The goal of this process was to give an unbiased sample of the current state of scientific V̇O~2max~ testing. Based on the procedure described in the preregistration, the sample included a total of 500 articles.

The abstracts from the articles included in the random sample were blinded for scanning. This meant the removal of any further information not relevant for the screening---such as authors or journals---leaving only the title, abstract and an ID of the article (see Appendix C for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in Table 3. When the screeners disagreed in their assessments, they resolved conflicts by discussion.

```{r exclusion, echo = FALSE}
exclusion_criteria <- data.frame(
  id = c("A*", "B*", "C*", "D*", "E", "F"),
  criteria = c("not in English", "no primary research", "research not in humans", "VO2max only estimated", "no appropriate test protocol", "no information regarding the exclusion criteria"),
  details = c("Full text only available in non-English language", "research was no original investigation or only a reanalysis of data", "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", "Protocol for VO2max testing did either not include exercise to voluntial exhaustion or was to long (>20 min) for a reliable estimate", "crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing")
)

knitr::kable(
  exclusion_criteria,
  col.names = c("", "Criterium", "Details"),
  caption = "Exclusion criteria for the screening process. During the abstract screening only the criteria marked with an asterix were evaluated, during full text screening all were assessed."
)
```

After the abstract screening I retrieved the full texts for all articles remaining in the review. The full texts were again independently scanned by two researchers to include only those articles that measured V̇O~2max~ using an appropriate testing procedure in humans (see Table 3 for the detailed full-text exclusion criteria). Conflicts were resolved by discussion.

All data exclusion steps are documented in an Markdown script on GitHub (LINKXXX).

@@

The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstract scanning, thus not stated in the preregistration. All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).

@@

The extracted data is presented in a purely descriptive way. I calculated the relative and absolute frequency for the reporting of the extracted items. Similarly, I brook down the different data strategies use for processing data in all articles that reported the use of breath-by-breath measurements. Total interval duration of averaging procedures were derived from the reported parameters.

@@

To determine the influence the most common data processing strategies have on the determination of V̇O~2max~, I compared them on a set of already collected gas exchange data from running.

@@

A total of N = 72 exercise test were analysed for this study. Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests. The data was from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (15 female, 55 male; three of the males participated in both studies). The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b] using exactly the same exercise protocol. Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for ten minutes at a velocity of 2.8 m/s as a warm-up. After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m/s for two minutes and subsequently increased velocity by 0.15 m/s every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

Gas exchange data were recorded using a ZAN 600 device (nSpire Health, Inc., Longmont, CO, United States of America). The device was calibrated with a 3l-syringe pump (nSpire Health, Inc., Longmont, CO, 143 United States of America) and a reference gas (15% O~2~, 6% CO~2~) before each measurement. The measured breath-by-breath data is available on GitHub (LINKXXX).

@@

To compare different processing methods within and between individuals, I choose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level I calculated the median and 10%- and 90%-quantiles of each processing strategy.

In an additional exploratory I investigated the respiratory rate (number of breathes per minute) during the ramp tests. This may help to understand how breath-based and time-based data processing methods relate to each other.

### 2022-05-02 thesis/results.Rmd
Continue writing thesis

Reporting practices of the methodology of gas exchange measures differed widely within the literature (see Table 4). More than half (51.8%) of the articles did not report any information on their data processing strategy. Only one in twenty articles (5.4%) provided a rationale for their used strategy.

@@

@@

![VO2max varies by data processing strategy. Values are expressed relative to the VO2max from a 30-second binned average --- the most common strategy as determined by the review. Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile. Using moving average leads to systematically higher VO2max values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to changes in VO2max as large as 5%.](../plots/comparison.png)

![Respiratory Rates peak around 60min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72). .](../plots/rr.png)

>>>>

Reporting practices of the methodology of gas exchange measures differed widely across the literature (see Table 4). More than half (51.8%) of the articles did not report any information on their data processing strategy. One in twenty articles (5.4%) provided a rationale for their used strategy. Only a single article [@maturana2021] reported information regarding all the investigated criteria.

@@

Data preprocessing

@@

The average V̇O~2max~ as determined by a binned 30-second average was 62.2±6.3 (mean ± standard deviation). Applying different data processing strategies for V̇O~2max~ determination leads to different outcome values (see Figure 4).

![V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average --- the most common strategy as determined by the review. Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%,.](../plots/comparison.png)

Binned time averages systematically lead to lower V̇O~2max~ values compared to moving averages. Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~-values. Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.

Moving time and moving breath averages yielded nearly identical values for V̇O~2max~ over all calculation intervals. This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see Figure 4).

![Respiratory Rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72).](../plots/rr.png)
